---
title: 'R Notebook'
output: html_notebook
---
# Introductory R Hands-On Workshop
## Variables & Data Types
### Part 1.1 - VARIABLES: CHARACTERS & INTEGERS 
Create a new script: File -> New file -> R Script
Now let’s create a variable! 
```{r}
# Assign variable a the value 2. You can use for that the assignment operator: either <- or =
a <- 2
a
```


```{r}
# Unlike other languages variables can contain dots 
a.1 = 3
a.1
```

Look at the Environment section to see variables you’ve made and their current values.
Change the value of a, then look at the Environment section to see the updated value.
```{r}
a <- 3 + 5
a
```

c() creates a vector: a sequence of data points of the same type
```{r}
b <- c(1,3,5,7,9)
b
```

```{r}
b.1 <- 1:12
b.1
```
```{r}
b.2 <- sequence(12)
b.2
```

Now try this: what happens?
```{r}
c <- c(4,3,'5',7)	
c
```

Let’s explore. Try this: 
```{r}
c <- '5'
c
```

```{r}
# Now try this:
c <- 5
c
```

```{r}
a <- 'RStudio'		# quotation marks make character variables
a.3 <- as.character(5)	# this is another way to make the character variable type

a
a.3
```

### Part 1.2 - VARIABLES: BOOLEAN DATA

Boolean or logical variables only have true and false options. Computers see “true” as 1 and “false” as 0.
Make this variable and then look at the Environment section to look at the variable type.

```{r}
a.5 <- TRUE
a.6 <- c(TRUE, FALSE, FALSE, TRUE)

a.5
a.6
```

There are also questions you can ask R that will give boolean answers. Ask R whether a is a numeric (non-character) variable.

```{r}
is.numeric(a)
```

A: a is a boolean variable, not a numeric variable. Even though the computer sees true as 1 and false as 0, it can tell that you made a boolean variable because you typed TRUE and FALSE when you created it. 

Now try these questions:
```{r}
is.logical(a); is.character(a)
```

### Part 2 - FUNCTIONS
We have been using functions to ask R questions and to give it tasks. The functions we have used so far are:
```{r}
c()
as.character(5)
is.numeric(a)
is.logical(a)
is.character(a)
```

Functions have a function name (c, as.character, is.numeric etc). The function name is followed by curved brackets:()
Functions always have these brackets, even if there’s nothing inside. The area inside the brackets can be used to give the function information
```{r}
as.character(5)
```

Here we are giving the function as.character() the information: 5. 
This means that it checks whether 5 is a character. If we wanted to store the result that the function gives us, we can make the answer into a stored variable like this:
```{r}
b <- as.character(5)
```

Q1 --> What variable type would b be, and what would its value be? 
A: 

### Part 3 - DATA FRAMES: USED FOR GRAPHING AND STATISTICAL ANALYSES

The data.frame() function makes data frames out of your data. A data frame is like a table: it stores your data neatly. The data frame structure is used in most applications of R.
```{r}
# Create three vectors: a, b, c.
# (Reminder: a vector is a sequence of data that has the same variable type)
a <- c(1,2,3,4,10,11,12)
b <- c(5,6,7,8,13,14,15)
c <- c('yes','no','no','yes','no','yes','yes')
```

Combine these vectors to make and store a data frame called myData (or call it something else if you want!)
```{r}
myData <- data.frame(a,b,c)

# Now run this function:
head(myData,3)
```

#### ACCESSING DATAFRAMES ELEMENTS

Data frames use the [ROW, COLUMN] access structure:

Example 1: To access the element in the 4th row, 2nd column of the data frame:
```{r}
myData[4,2]
```

```{r}
# Accessing COLUMNS - To access a whole column of the data frame, use the operator '$'. This allows you to find information by name. let’s find all the values with column name a:
myData$a
```

```{r}
# Column
myData[,1]
```

```{r}
# Row
myData[1,]
```


## Practical Exercises

#### STEP 1: Set the working directory

You can import data into R from lots of different file types. You can also import it from any location.
Today we’re going to import data from a file you have on your computer.

1. The file is called 'diamonds.csv', you should have downloaded it (if not, time to download now!). Lookup link for some documentation: https://www.kaggle.com/shivam2503/diamonds
2. Find out where your diamonds.csv file is: probably in your /Downloads folder.
3. Check where Rstudio looks to find new data.
4. “Working directory” is the name that Rstudio uses for the place its data is stored. To find out, enter this line (get working directory):
```{r}
getwd()
```

This function asks Rstudio to tell you where the current working directory is. This is probably not set to your Downloads folder at the moment. To set it to your Downloads folder, type:
```{r}
setwd("~/Documents/GitHub/meetup_presentations_miami/20170823_intro_workshop/")
# The ~ fills in all the other file location information automatically
```

Now enter this line again to check that you’re in the same location as your diamonds.csv file:
```{r}
getwd()
```

### STEP 2: IMPORTING DATA AND EXPLORING DATA

1. Load your data
2. Read data set using a function that loads csv files
```{r}
library(readr)
diamonds <- read_csv("~/Documents/GitHub/meetup_presentations_miami/20170823_intro_workshop/diamonds.csv")
# Dataset consists of the price and quality information for 54,000 diamonds
```

To have a quick look at the first few rows of the dataset, type:
```{r}
head(diamonds)
```

You can also have a quick look at the last few rows
```{r}
tail(diamonds)
```

To see how RStudio stores the diamonds data set, print the structure:
```{r}
str(diamonds)
```

Q2 --> How many variables and observations does the data contain?
Q3 --> What types of variables does the data contain?

#### A Summary of Variable Types:
* numeric = any number eg. 1 or 5.43
* int = integer = discrete value with a numerical component eg. 2L, 3L, 4L etc
* bool = boolean (also known as logical) = TRUE or FALSE value 
* char = character = read as a word eg 'hello'
* factor = discrete value eg. 'good', 'better', 'best'


Another helpful function is summary:
```{r}
summary(diamonds)
```

### STEP 3: GRAPHING THE DATA
R is really useful because of its packages

*What’s a package?*
A package is some code that someone else has taken the time to perfect. Each package is usually useful for one particular application.
For example, there is a package called 'dplyr' that is useful for data sorting and filtering


(...Continue with the exercise)
We want to make beautiful graphs, and to be able to customise them easily. A great package for this is ggplot2.

#### Install & Load package
```{r}
install.packages("ggplot2")

# load package 
library(ggplot2)
```

### GRAPHING DATA WITH GGPLOT(): SCATTERPLOT
The package ggplot2 includes a function called ggplot(). We use the ggplot() function to create graphs using our data.

Q4 --> Let’s see how diamond price changes due to carat.
```{r}
ggplot(data = diamonds, 
       aes(x = carat, y = price)) + geom_point()
```

Let's deconstruct this command:

* ggplot – function that creates the basic ggplot graph
* data - the data frame that you want to use 
* aes – short for aesthetic, describes how your variables are graphed
* geom – the functions that tell ggplot how you want the data presented (scatter plot, histogram etc)

### GRAPHING DATA WITH GGPLOT(): SCATTERPLOT WITH DETAILS

Lets add some colour (American spelling is generally used in R, so it’s color!)
```{r}
ggplot(data = diamonds, 
       aes(x = carat, 
           y = price, 
           color = cut)) + geom_point()
```

Let’s make it more readable:  
```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) + geom_point() +
 facet_wrap(~cut) + ggtitle("Diamonds price vs carat")
```

Lets explore the price range of diamonds
```{r}
ggplot(data = diamonds, aes(x = price)) + geom_histogram()
```

Let’s make it better!
```{r}
ggplot(data = diamonds, aes(x = price)) + 
geom_histogram(binwidth = 500, col = "black", fill = "lightyellow") +
ggtitle("histogram of diamond prices") +
xlab("price, $")
```

### Try by yourself:

1. Experiment with the binwidth value.
2. Use scale_x_continuous() to see what happens:
```{r}
ggplot(data = diamonds, aes(x = price)) + 
geom_histogram(binwidth = 500, col = "black", fill = "lightyellow") +
scale_x_continuous(limits = c(0,5000))+ 
ggtitle("Histogram of diamond prices")+
xlab("price, $")
```

Q5 --> What does this histogram tell you?

### GRAPHING DATA WITH GGPLOT(): BOXPLOT

Let's explore boxplots:
```{r}
ggplot(data = diamonds, 
       aes(x = cut, y = price)) + geom_boxplot()
```

GRAPHING DATA WITH GGPLOT(): BOXPLOT WITH DETAILS
```{r}
ggplot(data = diamonds, aes(x = cut, y = price)) + 
geom_boxplot(color = "#1F3552", fill = "#4271AE", alpha = 0.7, outlier.shape = 1)+ theme_classic()+ scale_y_continuous(name = "diamond price, $", breaks = seq(0, 20000, 2500),limits=c(0, 20000)) + scale_x_discrete(name = "diamond cut") + ggtitle("Boxplot of diamond price by cut")
```

## RESOURCES:
* Stack Overflow: https://stackoverflow.com/documentation/r/topics
* Free Udacity ‘Data Analysis with R’ course: https://www.udacity.com/course/data-analysis-with-r--ud651
* Free Coursera ‘R Programming’ course: https://www.coursera.org/learn/r-programming
* R for Data Science by G.Grolemund and H. Wickham: http://r4ds.had.co.nz/index.html

## RStudio Help
Use R internal help systems by typing directly into the window:
?read.csv()
?mean()
help()






